Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example

### 2023-03-25 manuscript/main.qmd
Work in progress...

Reconstructing Earth's past climate is fundamental to understanding how ecosystems respond to climatic and environmental change. To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023]. Geochemical proxies such as oxygen isotopes (𝛅^18^O) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions. However, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

@@

To address these challenges, we present a novel Bayesian model that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.

Routinely, one of three approaches are used to reconstruct past conditions one of three approaches

@@

Here, we present a Bayesian, hierarchical model for inferring the latitudinal gradient of sea surface temperatures (SSTs) that expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data.

@@

To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle \[\@...\]. As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq. ? and ?. To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples. This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition. The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for a given sample size (5, 10, 20, and 50). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse (flat), greenhouse (quadratic), icehouse (linear), extreme icehouse (steep). The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs. The blue shading depicts the 90%, 95%, and 99% credible intervals. Each gradient is depicted in absolute latitude.

:::

>>>>

Reconstructing past climate is fundamental to understanding the long-term evolution of the Earth's climate system, including natural climate variability [@zachos2008], the impacts of external forcings such as impact events [@chiarenza2020], and changes in greenhouse gas concentrations [@foster2017]. Moreover, palaeoclimatic reconstructions are vital for contextualising current global warming, enabling evaluation of the magnitude and significance in a broader historical perspective [@royer2004; @Burke2018], as well as understanding ecosystem response to environmental change [@jones2022reef; @woodhouse2023].

Recent advances in deep-time climate reconstructions [e.g. @Burgener2023; @chevalier2022], including improvements in proxy data quantification [e.g. @bao2023] and data synthesis [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022], are enabling researchers to better understand the complex and dynamic nature of the Earth's climate system over geologic time. However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially biased, impacted by incomplete preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022]. Recent work has demonstrated that this bias can lead to spurious estimates of latitudinal temperature gradients and global temperatures [@judd2020; @Jones2022].

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data.

----

how ecosystems respond to climatic and environmental change. To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023]. Geochemical proxies such as oxygen isotopes (𝛅^18^O) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions. However, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

@@

@@

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF). We then randomly sampled (1000 iterations) these gradients using a decreasing sample size (50, 20, 10, and 5) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types (**Should we be using some kind of statistic here to say that the curves are similar, perhaps the Kolmogorov-Smirnoff (K-S) Statistic?**). Subsequently, we test whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples, providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.

To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples. This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition. The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).

we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle \[\@...\]. As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq. ? and ?. To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples. This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition. The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for a given sample size (5, 10, 20, and 50). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse (flat), greenhouse (quadratic), icehouse (linear), extreme icehouse (steep). The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs. The blue shading depicts the 90%, 95%, and 99% credible intervals. Each gradient is depicted in absolute latitude.
:::
